Testing Quasar Unification: I. Radiative Transfer in Clumpy Winds

### 2016-01-21 paper.tex
modding some wording for the reviewer

Our approach to line transfer is based upon the macro-atom implementation developed by 
\cite{lucy2002, lucy2003}, in which the energy flows through the system are described in 
terms of indivisible energy quanta of radiant or kinetic energy 
(`$r-$packets' and `$k-$packets' respectively; see also section~\ref{sec:photon_sources}),
which are reprocessed by `macro-atoms' with associated transition probabilities.
We use the Sobolev approximation \citep[e.g.][]{sobolev1957,sobolev1960,rybickihummer1978}
to compute the location of line interactions.
In our case, for reasons of computational efficiency, we adopt the  hybrid macro-atom scheme 
described by M15. In this scheme, the energy packets interact with either two-level 
`simple atoms' or full macro-atoms. 
This allows one to treat non-LTE line transfer in radiative equilibrium
without approximation for elements that are identified as 
full macro-atoms, while maintaining the fast `two-level' 
treatment of resonance lines when elements are identified 
as simple atoms (see M15). In this study,
only H and He are treated as macro-atoms, because the process is computationally
intensive, and we expect recombination to be particularly important
in determining their level populations and resultant line emission.
We are also especially interested in the contribution to 
AGN spectra of \LA.  H13 treated all atoms in a two-level approximation.

@@

Macro-atoms have their ion and level populations derived from
MC rate estimators as described by Lucy (2002,2003). 
Previously (LK02, H13, M15), we used a modified Saha approach to 
calculate the ionization fractions
of simple-ions. As part of  this effort, we have 
now improved {\sc PYTHON} to explicitly solve the 
rate equations between ions in non-LTE. This dispenses with a number of small assumptions 
made in the modified Saha approach, is more numerically stable, and
prepares for a more accurate treatment of other physical processes in future. 
% in principle, allows 
% the direct addition of extra physical processes that would previously have necessitated 
% approximate treatments.

In order to calculate the photoionization rate, 
we model the SED in a grid cell using the technique described by H13. In this scheme,
the mean intensity, $J_{\nu}$ in a series of $n$ bands is modeled as either a power law or exponential
in frequency $\nu$, with the fit parameters deduced from band-limited radiation field estimators.
This allows the calculation of a photoionization rate estimator. Ion abundances are
then calculated by solving the rate equations between ions. We include collisional ionization and photoionization balanced with radiative, 
dielectronic and collisional (three-body) recombination.
As in M15, we use a dilute Boltzmann approximation to calculate 
the populations of levels for simple-ions. This dilute approximation 
is not required for macro-atom levels.

@@

We include  free-free, bound-free and bound-bound heating
and cooling processes in the model. For radiative transfer purposes
we treat electron scattering in the Thomson limit, 
but take full account of Compton heating and cooling when
calculating the thermal balance of the plasma (see H13).
Adiabatic cooling is included, but is insignificant 
in most of the outflow.

@@

Our kinematic prescription for a biconical disc wind model
follows \cite{SV93}, and is described further by
LK02, H13 and M15. A schematic is shown in Fig.~\ref{fig:cartoon},
with key aspects marked. The general biconical
geometry is similar to that invoked by \cite{MCGV95} and 
\cite{elvis2000} to explain the phenomenonology
of quasars and BALQSOs.

@@

In the SV93 model, a biconical disc wind rises from the accretion 
disc between launch radii $r_{min}$ and $r_{max}$.
The opening angles of the wind are set to $\theta_{min}$ and $\theta_{max}$.
The poloidal velocity along each individual streamline at a poloidal distance $l$ 
is then given by
\begin{equation}
v_l=v_0+\left[v_{\infty}(r_0)-v_0\right]\frac{\left(l/R_v\right)^{\alpha}}{\left(l/R_v\right)^{\alpha}+1},
\label{v_law}
\end{equation}
where $v_0$ is the velocity at the base of the streamline, $\alpha$ is
an exponent governing how quickly the wind accelerations and 
$R_v$ is the `acceleration length', defined as the distance at which
the outflow reaches half of its terminal velocity, $v_{\infty}$.
The terminal velocity is set to a fixed multiple of the escape
velocity, $v_{esc}$, at the base of the streamline (radius $r_0$).
The rotational velocity, $v_{\phi}$, is initially Keplerian ($v_k = [GM/r_0]^{1/2}$),
and the wind conserves specific angular momentum, such that 
\begin{equation}
v_{\phi} r = v_k r_0.
\label{v_law}
\end{equation}
The velocity law is crucial in determining the output spectra,
as it affects not only the projected velocities along the line of sight,
but also the density and ionization state of the outflow.
A wind that accelerates more slowly will have a denser wind base
with correspondingly different ionization and emission characteristics.

@@

\begin{itemize}
\item UV absorption lines 
with $BI > 0$ at $\sim20\%$ of viewing angles (e.g. Knigge et al. 2010);
\item Line emission emerging at low inclinations, with $EW\sim40$\AA in \civline\ \citep[e.g. ][]{shen2011};
\item H recombination lines with $EW\sim50$\AA\ in \la\ \citep[e.g. ][]{shen2011};
\item  \mg\ and \al\ (LoBAL) absorption features with $BI > 0$ at a subset of 
BAL viewing angles;
\item Verisimilitude with quasar composite spectra.
\end{itemize}
Here $BI$ is the `Balnicity Index' (Weymann et al. 1991), given by
\begin{equation}
BI = \int^{25000~{\rm km~s}^{-1}}_{3000~{\rm km~s}^{-1}} \left( 1 - \frac{f(v)}{0.9} \right) dv.
\end{equation}
The constant $C=0$ everywhere, unless the normalized flux
has satisfied $f(v)<0.9$ continuously for at least $2000$~km~s$^{−1}$, 
whereby $C$ is set to $1$.

@@

\begin{figure} 
\centering
\includegraphics[width=0.5\textwidth]{figures/fig4.eps}
\caption
{
A cartoon describing the broad classes of sightline 
in the fiducial model, illustrating how geometric effects lead to 
the different emergent spectra. The colour gradient is approximate,
but indicates the stratified ionization structure, 
from highly ionized (yellow) to low ionization (purple) material.
%[FIGURE NEEDS IMPROVING]
}
\label{fig:sightline}
\end{figure}

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/fig5.eps}
\caption
{
Synthetic spectra at two viewing angles, 
this time in frequency space and including the optical band,
compared to the non-BAL SDSS quasar composite. The spectra are normalised to the flux at 
$2000$\AA, then an offset of 2 is applied per spectrum for clarity -- the dotted lines show the zero point of $F_\nu / F_{2000\AA}$ in each case.
}
\label{fig:sed}
\end{figure*}

@@

In Fig.~\ref{fig:sed}, we show an $F_{\nu}$ spectrum with broader waveband coverage
that includes the optical, showing that our synthetic spectra 
also exhibit \ha\ and \hb\ emission. 
In this panel, we include a low inclination and 
also a very high inclination 
spectrum, which looks underneath the wind cone. This model shows 
strong line emission with very similar widths and line ratios to the quasar composites, and
the Balmer lines are double peaked, due to velocity projection effects.  
Such double-peaked lines are seen in so-called `disc emitter' systems 
\citep[e.g.][]{eracleous1994} but not the majority of AGN.     
The line equivalent widths (EWs) increase at high inclination
due to a weakened continuum from wind attenuation, 
disc foreshortening and limb darkening. This effect also 
leads to a redder continuum slope, as seen in quasars, which is
due to Balmer continuum and Balmer and Fe~\textsc{ii} line emission.
This extreme $89^\circ$ viewing angle cannot represent a typical quasar within a unified model,
but does show that a model such as this can naturally reproduce quasar emission lines
if the emissivity of the wind is increased {\em with respect to the disc continuum}.
In addition, it neatly demonstrates how a stratified outflow can naturally
reproduce the range of ionization states seen in quasars.

@@

We have carried out MCRT simulations using a simple
prescription for a biconical disc wind, with
the aim of expanding on the work of H13. To do this, we introduced two main
improvements: First, we included a simple treatment of clumping, and second, 
we improved the modelling of recombination lines by treating H and He as
`macro-atoms'. Having selected a fiducial model from an initial simulation grid,
we assessed the viability of such a model for geometric 
unification of quasars, and found the following main points:
\begin{enumerate}
\item Clumping moderates the ionization state
sufficiently to allow for the 
formation of strong UV BALs while agreeing well with the X-ray
properties of luminous AGN and quasars. 
\smallskip
\item A clumpy outflow model naturally 
reproduces the range of ionization states
expected in quasars, due to its stratified density
and temperature structure. 
LoBAL line profiles are seen at a subset of viewing angles, and Fe~\textsc{ii}
absorption is seen at particularly high inclinations. 
\smallskip
\item The synthetic spectra show \la\ line and weak He~\textsc{ii}~$1640$\AA\ line
as a result of our improved treatment of recombination using macro-atoms. We also see
Balmer emission lines and a Balmer recombination continuum in the optical spectrum, but this
is only really significant at high inclination where to continuum is suppressed.  
\smallskip
\item The higher X-ray luminosity causes a significant 
increase in the strength of the collisionally excited emission
lines produced by the model. 
However, the equivalent-width ratios of the emission lines do not match
observations, suggesting that a greater volume of dense ($n_e\sim10^{10}$~cm$^{-3}$)
material may be required.
\smallskip
\item The line EWs in the synthetic spectra increase with inclination.
BAL and non-BAL quasar composites have comparable EWs, so our model
fails to reproduce this behaviour.
 If the BLR emits fairly isotropically then for a 
foreshortened, limb-darkened accretion disc 
it is not possible to achieve line ratios at low inclinations 
that are comparable to those at high inclinations. 
We suggest that understanding the angular distribution of 
line and continuum emission is a crucial question for theoretical models.
% Furthermore, this is conclusion is independent of the assumed 
% BLR geometry and size.
\end{enumerate}
Our work confirms a number of expected outcomes from a geometric unification 
model, and suggests that a simple biconical geometry such as this can come close to 
explaining much of the  phenomenology of quasars. However, our conclusions pose 
some challenges to a picture in which BALQSOs are
explained by an {\em equatorial} wind rising from a classical thin disc, and suggest 
the angular distribution of emission is important to understand if this 
geometry is to be refuted or confirmed. We suggest that obtaining reliable 
observational orientation indicators and 
exploring a wider parameter space of outflow geometries in simulations
are obvious avenues for future work.

@@

The work of JHM, SWM, NSH and CK is supported by the
Science and Technology Facilities Council (STFC),
via two studentships and a consolidated grant, respectively.
CK also acknowledges a Leverhulme fellowship.
We would like to thank Omer Blaes, Ivan Hubeny and Shane Davis for their
assistance with \agn. We are grateful to Mike Brotherton, Mike DiPompeo,
Sebastien Hoenig and Frederic Marin for helpful correspondence regarding
polarisation measurements and orientation indicators.
We would also like to thank Daniel Proga, Daniel Capellupo, Sam Connolly and
Dirk Grupe for useful discussions.  Simulations were conducted using \py\ version 80,
and made use of the IRIDIS High Performance Computing Facility at the
University of Southampton. Figures were produced using the {\tt matplotlib} plotting library \citep{matplotlib}. This work made use of the Sloan Digital Sky Survey.
Funding for the Sloan Digital Sky Survey has been provided by
the Alfred P. Sloan Foundation, the U.S. Department of Energy Office of
Science, and the Participating Institutions.

>>>>

Our approach to line transfer is based upon the macro-atom implementation developed by 
\cite{lucy2002, lucy2003}, in which the energy flows through the system are described in 
terms of indivisible energy quanta of radiant or kinetic energy 
(`$r-$packets' and `$k-$packets' respectively; see also section~\ref{sec:photon_sources}),
which are reprocessed by `macro-atoms' with associated transition probabilities.
{\bf We use the Sobolev approximation \citep[e.g.][]{sobolev1957,sobolev1960,rybickihummer1978}
to compute the location and optical depth of line interactions}.
In our case, for reasons of computational efficiency, we adopt the  hybrid macro-atom scheme 
described by M15. In this scheme, the energy packets interact with either two-level 
`simple atoms' or full macro-atoms. 
This allows one to treat non-LTE line transfer in radiative equilibrium
without approximation for elements that are identified as 
full macro-atoms, while maintaining the fast `two-level' 
treatment of resonance lines when elements are identified 
as simple atoms (see M15). In this study,
only H and He are treated as macro-atoms, because the process is computationally
intensive, and we expect recombination to be particularly important
in determining their level populations and resultant line emission.
We are also especially interested in the contribution to 
AGN spectra of \LA.  H13 treated all atoms in a two-level approximation.

@@

Macro-atoms have their ion and level populations derived from
MC rate estimators as described by Lucy (2002,2003). 
Previously (LK02, H13, M15), we used a modified Saha approach to 
calculate the ionization fractions
of simple-ions. As part of  this effort, we have 
now improved {\sc PYTHON} to explicitly solve the 
rate equations between ions in non-LTE. This dispenses with a number of small assumptions 
made in the modified Saha approach, is more numerically stable, and
prepares for a more accurate treatment of other physical processes in future.

In order to calculate the photoionization rate, 
we model the SED in a grid cell using the technique described by H13. In this scheme,
the mean intensity, $J_{\nu}$, in a series of $n$ bands is modeled as either a power law or exponential
in frequency $\nu$, with the fit parameters deduced from band-limited radiation field estimators.
This allows the calculation of a photoionization rate estimator. Ion abundances are
then calculated by solving the rate equations between ions. We include collisional ionization and photoionization balanced with radiative, 
dielectronic and collisional (three-body) recombination.
As in M15, we use a dilute Boltzmann approximation to calculate 
the populations of levels for simple-ions. This dilute approximation 
is not required for macro-atom levels.

@@

We include  free-free, bound-free and bound-bound heating
and cooling processes in the model. For radiative transfer purposes
we treat electron scattering in the Thomson limit, 
but take full account of Compton heating and cooling when
calculating the thermal balance of the plasma (see H13).
{\bf 
Adiabatic cooling in the cell is calculated using the estimator
\begin{equation}
C_{ad} = f_V n_e k_B T_e V (\Delta \cdot v),
\end{equation}
where $V$ is the cell volume, $T_e$ is the electron temperature,
$n_e$ is the electron density and $\Delta \cdot v$ is the divergence of the 
velocity field at the centre of the cell.
For self-consistency, $k$-packets then have a finite 
destruction probability of 
\begin{equation}
P_{k,destruct} = \frac{C_{ad}}{C_{tot}},
\end{equation}
where $C_{tot}$ is the sum over all cooling rates in the cell.
We find that adiabatic cooling is insignificant in much of
the outflow.

@@

Our kinematic prescription for a biconical disc wind model
follows \cite{SV93}, and is described further by
LK02, H13 and M15. A schematic is shown in Fig.~\ref{fig:cartoon},
with key aspects marked. The general biconical
geometry is similar to that invoked by \cite{MCGV95} and 
\cite{elvis2000} to explain the phenomenonology
of quasars and BALQSOs. {\bf However, we do not include the initial
vertical component present in the Elvis (2000) model, which was
invoked as a qualitative explanation for narrow absorption lines}.

@@

In the SV93 model, a biconical disc wind rises from the accretion 
disc between launch radii $r_{min}$ and $r_{max}$.
The opening angles of the wind are set to $\theta_{min}$ and $\theta_{max}$.
The poloidal velocity along each individual streamline at a poloidal distance $l$ 
is then given by
\begin{equation}
v_l=v_0+\left[v_{\infty}(r_0)-v_0\right]\frac{\left(l/R_v\right)^{\alpha}}{\left(l/R_v\right)^{\alpha}+1},
\label{v_law}
\end{equation}
where $v_0$ is the velocity at the base of the streamline, $\alpha$ is
an exponent governing how quickly the wind accelerations and 
$R_v$ is the `acceleration length', defined as the distance at which
the outflow reaches half of its terminal velocity, $v_{\infty}$.
The terminal velocity is set to a fixed multiple of the escape
velocity, $v_{esc}$, at the base of the streamline (radius $r_0$).
The rotational velocity, $v_{\phi}$, is initially Keplerian ($v_k = [GM/r_0]^{1/2}$),
and the wind conserves specific angular momentum, such that 
\begin{equation}
v_{\phi} r = v_k r_0.
\label{v_law}
\end{equation}
The velocity law is crucial in determining the output spectra,
as it affects not only the projected velocities along the line of sight,
but also the density and ionization state of the outflow.
A wind that accelerates more slowly will have a denser wind base
with correspondingly different ionization and emission characteristics.
}

@@

\begin{itemize}
\item UV absorption lines 
with $BI > 0$ at $\sim20\%$ of viewing angles (e.g. Knigge et al. 2010);
\item Line emission emerging at low inclinations, with $EW\sim40$\AA\ in \civline\ \citep[e.g. ][]{shen2011};
\item H recombination lines with $EW\sim50$\AA\ in \la\ \citep[e.g. ][]{shen2011};
\item  \mg\ and \al\ (LoBAL) absorption features with $BI > 0$ at a subset of 
BAL viewing angles;
\item Verisimilitude with quasar composite spectra.
\end{itemize}
Here $BI$ is the `Balnicity Index' (Weymann et al. 1991), given by
\begin{equation}
BI = \int^{25000~{\rm km~s}^{-1}}_{3000~{\rm km~s}^{-1}} \left( 1 - \frac{f(v)}{0.9} \right) dv.
\end{equation}
The constant $C=0$ everywhere, unless the normalized flux
has satisfied $f(v)<0.9$ continuously for at least $2000$~km~s$^{−1}$, 
whereby $C$ is set to $1$.

@@

\begin{figure} 
\centering
\includegraphics[width=0.5\textwidth]{figures/fig4.eps}
\caption
{
A cartoon describing the broad classes of sightline 
in the fiducial model, illustrating how geometric effects lead to 
the different emergent spectra. The colour gradient is approximate,
but indicates the stratified ionization structure, 
from highly ionized (yellow) to low ionization (purple) material.
}
\label{fig:sightline}
\end{figure}

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/fig5.eps}
\caption
{
Synthetic spectra at two viewing angles, 
this time in frequency space and including the optical band,
compared to the non-BAL SDSS quasar composite. The spectra are normalised to the flux at 
$2000$\AA, then an offset of 2 is applied per spectrum for clarity -- the dotted lines show the zero point of $F_\nu / F_{2000}$ in each case.
}
\label{fig:sed}
\end{figure*}

@@

In Fig.~\ref{fig:sed}, we show an $F_{\nu}$ spectrum with broader waveband coverage
that includes the optical, showing that our synthetic spectra 
also exhibit \ha\ and \hb\ emission. 
In this panel, we include a low inclination and 
also a very high inclination 
spectrum, which looks underneath the wind cone. This model shows 
strong line emission with very similar widths and line ratios to the quasar composites, and
the Balmer lines are double peaked, due to velocity projection effects.  
Such double-peaked lines are seen in so-called `disc emitter' systems 
\citep[e.g.][]{eracleous1994} but not the majority of AGN.     
The line equivalent widths (EWs) increase at high inclination
due to a weakened continuum from wind attenuation, 
disc foreshortening and limb darkening. This effect also 
leads to a redder continuum slope, as seen in quasars, which is
due to Balmer continuum and Balmer and Fe~\textsc{ii} line emission.

{\bf This extreme $89^\circ$ viewing angle cannot represent a typical quasar within a unified model, as
extreme inclinations should be extremely underrepresented in quasar samples due to the
compton-thick absorber or `Torus' expected at high inclinations ().}
However, this spectrum does show that a model such as this can 
naturally reproduce quasar emission lines if the emissivity of the wind is 
increased {\em with respect to the disc continuum}.
In addition, it neatly demonstrates how a stratified outflow can naturally
reproduce the range of ionization states seen in quasars.

@@

We have carried out MCRT simulations using a simple
prescription for a biconical disc wind, with
the aim of expanding on the work of H13. To do this, we introduced two main
improvements: First, we included a simple treatment of clumping, and second, 
we improved the modelling of recombination lines by treating H and He as
`macro-atoms'. Having selected a fiducial model from an initial simulation grid,
we assessed the viability of such a model for geometric 
unification of quasars, and found the following main points:
\begin{enumerate}
\item Clumping moderates the ionization state
sufficiently to allow for the 
formation of strong UV BALs while agreeing well with the X-ray
properties of luminous AGN and quasars. 
\smallskip
\item A clumpy outflow model naturally 
reproduces the range of ionization states
expected in quasars, due to its stratified density
and temperature structure. 
LoBAL line profiles are seen at a subset of viewing angles, and Fe~\textsc{ii}
absorption is seen at particularly high inclinations. 
\smallskip
\item The synthetic spectra show \la\ line and weak He~\textsc{ii}~$1640$\AA\ line
as a result of our improved treatment of recombination using macro-atoms. We also see
Balmer emission lines and a Balmer recombination continuum in the optical spectrum, but this
is only really significant at high inclination where to continuum is suppressed.  
\smallskip
\item The higher X-ray luminosity causes a significant 
increase in the strength of the collisionally excited emission
lines produced by the model. 
However, the equivalent-width ratios of the emission lines do not match
observations, suggesting that a greater volume of dense ($n_e\sim10^{10}$~cm$^{-3}$)
material may be required.
\smallskip
\item The line EWs in the synthetic spectra increase with inclination.
BAL and non-BAL quasar composites have comparable EWs, so our model
fails to reproduce this behaviour.
 If the BLR emits fairly isotropically then for a 
foreshortened, limb-darkened accretion disc 
it is not possible to achieve line ratios at low inclinations 
that are comparable to those at high inclinations. 
We suggest that understanding the angular distribution of 
line and continuum emission is a crucial question for theoretical models.
\end{enumerate}
Our work confirms a number of expected outcomes from a geometric unification 
model, and suggests that a simple biconical geometry such as this can come close to 
explaining much of the  phenomenology of quasars. However, our conclusions pose 
some challenges to a picture in which BALQSOs are
explained by an {\em equatorial} wind rising from a classical thin disc, and suggest 
the angular distribution of emission is important to understand if this 
geometry is to be refuted or confirmed. We suggest that obtaining reliable 
observational orientation indicators and 
exploring a wider parameter space of outflow geometries in simulations
are obvious avenues for future work.

@@

The work of JHM, SWM, NSH and CK is supported by the
Science and Technology Facilities Council (STFC),
via two studentships and a consolidated grant, respectively.
CK also acknowledges a Leverhulme fellowship.
{\bf We would like to thank the anonymous referee for an insightful
and constructive report.}
We would like to thank Omer Blaes, Ivan Hubeny and Shane Davis for their
assistance with \agn. We are grateful to Mike Brotherton, Mike DiPompeo,
Sebastien Hoenig and Frederic Marin for helpful correspondence regarding
polarisation measurements and orientation indicators.
We would also like to thank Daniel Proga, Daniel Capellupo, Sam Connolly and
Dirk Grupe for useful discussions.  Simulations were conducted using \py\ version 80,
and made use of the IRIDIS High Performance Computing Facility at the
University of Southampton. Figures were produced using the {\tt matplotlib} 
plotting library \citep{matplotlib}. This work made use of the Sloan Digital Sky Survey.
Funding for the Sloan Digital Sky Survey has been provided by
the Alfred P. Sloan Foundation, the U.S. Department of Energy Office of
Science, and the Participating Institutions.
